---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Laura Exar"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r Set up Session}
#1
#Checked working directory
library(here)
here()
#Loaded packages
library(tidyverse); library(agricolae); library(ggthemes); library(lubridate)

#Imported the data
Lake.chem.phys <- read.csv(
file=here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
stringsAsFactors = TRUE
)

#Set date columns to date objects
Lake.chem.phys$sampledate <- mdy(Lake.chem.phys$sampledate) #do I need to change other date columns like year?

#2
#Built and set a theme
my_theme <- theme_base() +
theme(
axis.line = element_line(
linewidth = 1,
colour = "black"),
plot.background = element_rect(
color='grey'
),
plot.title = element_text(
color='blue'
)
)

theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes.
Ha: Mean lake temperature recorded during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
#Wrangled the dataset
Lake.chem.phys <- Lake.chem.phys %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5
#Created a scatterplot of temperature by depth
Temp.by.Depth <- Lake.chem.phys %>%
  ggplot(aes(x=depth, y=temperature_C)) +
  geom_point(size=0.5) + 
  labs(title="Temperature by Depth",
  x="Depth",
  y="Temperature") + 
    geom_smooth(method=lm, se=FALSE) +
    ylim(0, 35)
Temp.by.Depth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure suggests that temperature decreases as depth increases. Overall, the points have a large distributed at shallower depths (from around 0 to 7), but the distribution gets smaller at deeper depths.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
#Performed a linear regression of temperature by depth
Temp.by.Depth.Reg <- lm(data = Lake.chem.phys, temperature_C ~ depth)
summary(Temp.by.Depth.Reg)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth.

> Answer: The R-squared is 0.7387, which means that 73.87% of the variability in temperature can be explained by changes in depth. There are 9726 degrees of freedom. The p-value is < 2.2e-16; because the p-value is less than 0.05, we can reject the null hypothesis and accept the alterantive hypothesis that the variability in temperature can be explained by changes in depth. For every 1 m change in depth, the temperature is expected to change 1.94621 degrees C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
#Performed a multiple regression of temperature by depth, year, and daynum
Lake.ALL <- lm(data=Lake.chem.phys, temperature_C ~ year4 + daynum + depth)
summary(Lake.ALL)
step(Lake.ALL)

#10
#Ran multiple regression of temperature by depth
Lake.best <- lm(data=Lake.chem.phys, temperature_C ~ year4 + daynum + depth)
summary(Lake.best)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: Year4, daynum, and depth are the final set of explanatory variables that the AIC method suggests we use. The R-squared is 0.7411, which means that 74.11% of the variability in temperature can be explained by changes in depth. It is similiar, but a slight improvement, to the model using only depth as the explantory variable, because this model also only uses more explanatory variable that are all significant.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#Ran an ANOVA model to text whether the different lakes have different 
#temperatures
Lake.Temps.anova <- aov(data = Lake.chem.phys, temperature_C ~ lakename)
summary(Lake.Temps.anova)

#Ran a linear model to text whether the different lakes have different 
#temperatures
Lake.Temps.anova2 <- lm(data = Lake.chem.phys, temperature_C ~ lakename)
summary(Lake.Temps.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, because the models how a p-value of < 2.2e-16, which is less than 0.05, we reject the null hypthesis and accept the alternative hypothesis: there is a significant difference in mean temperature among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
#Created a scatterplot that shows temperature by depth
Temp.by.Depth.Lake <- Lake.chem.phys %>%
  ggplot(aes(x=depth, y=temperature_C, color= lakename)) +
  geom_point(alpha=0.5) + #alpha sets transparency
  labs(title="Temperature by Depth",
       x="Depth",
       y="Temperature", 
       color="")+
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0,35)
Temp.by.Depth.Lake

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
#Ran an Tukey HSD to determine which lakes have different means
TukeyHSD(Lake.Temps.anova)
Group <- HSD.test(Lake.Temps.anova, 'lakename', group=T)
Group

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake have the same mean temperature as Peter Lake, statistically speaking, as Peter Lake. None of the lakes have a mean temperature that is statistically distinct from all the other lakes.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we are just looking at Peter Lake and Paul Lake, we could use the two sample t-test to explore whether they have distinct mean temperatures. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#Wrangled the data to only include Crampton Lake and Ward Lake
Lake.chem.phys.2Lakes <- Lake.chem.phys %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

#Ran a two sample t-test to determine if the lake's July temperatures are 
#the same or different
Lake.twosample <- t.test(Lake.chem.phys.2Lakes$temperature_C ~ Lake.chem.phys.2Lakes$lakename)
Lake.twosample

```

>Answer: The test says that the mean July temperature for Crampton Lake is 15.35189 C, and the mean July temperature for Ward Lake is 14.45862 C. Because the p-value > 0.05, the mean temperatures for the lakes are not statistically significantly distinct. This confirms my results for part 16.